Proving Logical Atomicity using Lock Invariants\thanks{Supported by the National Science Foundation via grant \#1811894.}

### 2022-05-18 concurrency/ASL 2022/Verification of Concurrent Binary Search Tree.tex
Some fiddling to make the paper fit into 15 pages

Later, the TaDA logic~\cite{tada}, which codified the notion of logical atomicity, presented an alternative set of lock specifications:
\[\langle b.\ (\mathsf{L}(\ell) \land \neg b) \vee (\mathsf{U}(\ell) \land b) \rangle\ \texttt{acquire}(\ell)\ \langle \mathsf{L}(\ell) \land b \rangle \qquad \langle \mathsf{L}(\ell) \rangle\ \texttt{release}(\ell)\ \langle \mathsf{U}(\ell) \rangle\]
An \texttt{acquire} operation may access the lock repeatedly, finding it in either the locked ($\mathsf{L}$) or unlocked ($\mathsf{U}$) state; if it sees the unlocked state ($b$ holds), then it can atomically move to a locked state and return. A release simply moves atomically from the locked state to the unlocked state. These operations make no mention of a protected resource, but because they are atomic, they can interact with any global invariant (``shared region'' in TaDA's terminology): in particular, in combination with the invariant $(\mathsf{U}(\ell) * R) \vee \mathsf{L}(\ell)$, they can be used to derive the lock-invariant-based specs. It is also worth noting that these lock specifications are not axioms: they are proved for a simple spinlock implementation that uses atomic operations (e.g. compare-and-swap), which are the actual primitive operations in TaDA and the logics that follow it. In this style, tracking ownership of a lock is orthogonal to the lock assertion itself; a mechanism such as cancelable invariants~\cite{rustbelt-relaxed} can be used to both make the lock assertion $\mathsf{U}$/$\mathsf{L}$ publicly available and deallocate it once all threads are finished.

@@

Our proof technique is developed in a combination of two Coq-based separation logic verification systems, the Verified Software Toolchain (VST) and Iris. VST~\cite{plfcc} is a Coq-based system for proving separation logic specifications of C programs. It is proved sound against not only the C semantics of the CompCert compiler, but also the compiler correctness theorem~\cite{cpm}, so that verified programs are guaranteed to behave correctly when compiled and run. VST directly implements the concurrent separation logic of Hobor et al.~\cite{oraclesematic}, and the invariant-style lock specifications are incorporated into its soundness proof at a fundamental level.

@@

A thread-safe version of this data structure should satisfy atomic specifications that closely correspond to the sequential specs\footnote{We will also have functions to create and deallocate trees, but these functions need not be thread-safe, since the tree will not be shared when created and must not be shared when deallocated.}:
\begin{mathpar}
\langle m.\ \nodeboxrep_g\ p\ |\ \treerep_g\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep_g\ p\ |\ \treerep_g\ (m[k \mapsto v])\rangle

@@

\subsection{Atomicity with lock invariants}
The easiest way to synchronize a data structure is to add a lock that each operation acquires at the start and releases at the end. These \emph{coarse-grained} locking implementations are easy to verify with invariant-style lock specs: the lock's invariant can contain the whole data structure assertion (e.g. $\mathsf{BST}\:p\ t$), so that whichever thread is currently executing an operation on the tree at $p$ has full ownership of the data structure. If we only want to prove that the operation is thread-safe, we might choose as our lock invariant $\exists t.\ \mathsf{BST}\ p\ t$, asserting that we gain full ownership of the tree in some state $t$ every time we acquire the lock. On the other hand, if we want to prove specifications that describe changes to the data structure---as in the atomic specification $\langle m.\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \treerep\ (m[k \mapsto v])\rangle$---we need ghost state to track the changes.

@@

When an insertion encounters a node with key \lstinline{k}, it simply changes the value at that node to \lstinline{value}. If it reaches a leaf node, it allocates a new node with key \lstinline{k} and value \lstinline{v}, and inserts it at that position.
In both cases, we must show that we can atomically update the abstract state of the tree from $\treerep\ m$ to $\treerep (m[\texttt{k}\mapsto \texttt{v}])$, by showing that the update to the tree in memory corresponds to a local change in the abstract tree that inserts the new key-value pair. The proofs are split into a proof that the C code implements the local update, and a proof that the local update to a ghost-state tree implements the global update. In the first case, the local change is the insertion of $(\texttt{k}, \texttt{v})$ at an empty leaf whose range includes \lstinline{k}, and the addition of two new empty leaves as its children; in the second case, we simply change the value of an internal node whose key is already \lstinline{k}.

@@

\begin{figure}[htp]
\begin{subfigure}[t]{0.6\textwidth}
\begin{lstlisting}[language = C, basicstyle=\small\ttfamily, numbers=left]
void pushdown_left (tree_t *tgt){
  while(1) {
    tree *p = tgt->t;
    tree_t *rc = p->right;
    void *lq = rc->lock;
    acquire(rc->lock);
    tree *q = rc->t;
    if (q == NULL) {
      tree_t *lc = p->left;
      ... // move lc->t to tgt; deallocate p, lc, and rc
      release(tgt->lock);
      return;
    } else {
      turn_left(tgt, rc);
      tgt = q->left;
      release(tgt->lock);
    }
  }
}
\end{lstlisting} 
\end{subfigure}\qquad
\begin{subfigure}[t]{0.4\textwidth}
 \begin{lstlisting}[language = C, basicstyle=\small\ttfamily]

@@

The \lstinline{pushdown_left} function changes the structure of the tree in two ways. First, the rotation performed by \lstinline{turn_left} (line 14) changes the ranges of the node to be deleted and its right child (e.g., nodes 40 and 55 in the first step of figure~\ref{delete-diagram}); this is a local change, as the ranges of all other nodes (even the children of the rearranged nodes) are preserved. This is not a linearization point, so the rotation must be considered not to change the abstract state of the tree. We accounted for this by quantifying over the abstract tree $t$ in the definition of $\treerep$; rearranging nodes without changing the key-value map still implements the same abstract state, and other threads cannot assume that the tree structure they have observed is still the current structure of the tree. Second, the final removal of the node (line 10) increases the ranges of the subtree below it; for example, in the last step of figure~\ref{delete-diagram}, the range of node 35 changes from $(30, 40)$ to $(30, 50)$ when the node is removed. This is the linearization point of the function. The change occurs while \lstinline{pushdown_left} holds only the lock of the deleted node, so the algebra for range ghost state must allow a node's range to increase at any time, even when the node's lock is held. Fortunately, \lstinline{insert} and \lstinline{lookup} only use ranges to assert that the target key is in the range of the current node, which is still true if the range increases.

@@

\begin{figure}[htb]

@@

\section{Using the Specifications}

\begin{figure}[htp]
\begin{subfigure}[t]{0.4\textwidth}
\begin{lstlisting}[language = C, basicstyle=\small\ttfamily, numbers=left]
void* update_tree(void* t){
  a = 1;
  b = 2;
  c = 3; 
  d = 4;
  insert(t, 1, &a);
  insert(t, 2, &b);
  insert(t, 1, &c);
  insert(t, 2, &d);
  return NULL;
}
\end{lstlisting} 
\end{subfigure}\qquad
\begin{subfigure}[t]{0.55\textwidth}
 \begin{lstlisting}[language = C, basicstyle=\small\ttfamily]
int retrieve_value(treebox t){
  int y = -1;
  do {
     searchResult = lookup (t,  2);
     if (searchResult != NULL) {
        y = *((int *) searchResult);
     }
  } while (y != 4);
  void* searchResult = lookup (t,  1);
  return *((int *) searchResult);
}
\end{lstlisting}
\end{subfigure}
\caption{Client code}
\label{bst-conc-client}
\end{figure}

To demonstrate that the complexities of our lock-based proofs are hidden from prospective users, we verified a simple client program for the binary search tree, shown in figure \ref{bst-conc-client}. A producer thread running the \lstinline{update_tree} function populates a tree, while a consumer thread running \lstinline{retrieve_value} waits until key 2 is associated with a pointer to the value 4, and then returns the value associated with key 1. This is a data-structure-level version of the message-passing idiom, and demonstrates one of the advantages of atomic specifications: we can use them to synchronize threads and pass ownership of resources via data structure operations. The invariant for the BST \lstinline{t} is a simple state machine with three states, depending on the contents of the abstract map $m$: 1) $m(2) = \texttt{NULL}$, 2) $m(2) \mapsto i$ for some integer $i \neq 4$; and 3) $m(2) \mapsto 4 \land m(1) \mapsto 3$. In this final state, the invariant also holds the resources to be transferred from the producer to the consumer. We also include ghost state to reflect the fact that there is exactly one producer thread (so the producer always knows exactly which values are in the tree) and one consumer thread (so it can retrieve the resources without worrying about interference). The proof of the client uses only the atomic specifications of the BST operations, and is not affected by the style of lock specification.

@@

VeriFast~\cite{verifast}, a separation-logic-based verifier for C and Java programs, also supports lock-based and atomic concurrency~\cite{verifast-conc}, and has been used to verify a hand-over-hand-locking linked list similar to that of Krishna et al. The specifications of both the lock operations and the list itself use a precursor of TaDA's logical atomicity. VeriFast is not a foundationally verified system, but its basic logic is verified against the semantics of a toy language in Coq.

>>>>

Later, the TaDA logic~\cite{tada}, which codified the notion of logical atomicity, presented an alternative set of lock specifications:
\[\langle b.\ (\mathsf{L}(\ell) \land \neg b) \vee (\mathsf{U}(\ell) \land b) \rangle\ \texttt{acquire}(\ell)\ \langle \mathsf{L}(\ell) \land b \rangle \qquad \langle \mathsf{L}(\ell) \rangle\ \texttt{release}(\ell)\ \langle \mathsf{U}(\ell) \rangle\]
An \texttt{acquire} operation may access the lock repeatedly, finding it in either the locked ($\mathsf{L}$) or unlocked ($\mathsf{U}$) state; if it sees the unlocked state ($b$ holds), then it can atomically move to a locked state and return. A release simply moves atomically from the locked state to the unlocked state. These operations make no mention of a protected resource \LB{maybe add this:\emph{, i.e.~they obey the pattern of atomic specs for $P_l = Q_l = True$,} ?}, but because they are atomic, they can interact with any global invariant (``shared region'' in TaDA's terminology): in particular, in combination with the invariant $(\mathsf{U}(\ell) * R) \vee \mathsf{L}(\ell)$, they can be used to derive the lock-invariant-based specs. It is also worth noting that these lock specifications are not axioms: they are proved for a simple spinlock implementation that uses atomic operations (e.g. compare-and-swap), which are the actual primitive operations in TaDA and the logics that follow it. In this style, tracking ownership of a lock is orthogonal to the lock assertion itself; a mechanism such as cancelable invariants~\cite{rustbelt-relaxed} can be used to both make the lock assertion $\mathsf{U}$/$\mathsf{L}$ publicly available and deallocate it once all threads are finished.

@@

We employ a combination of two Coq-based verification systems, the Verified Software Toolchain (VST) and Iris. VST~\cite{plfcc} is a Coq-based system for proving separation logic specifications of C programs. It is proved sound against not only the C semantics of the CompCert compiler, but also the compiler correctness theorem~\cite{cpm}, so that verified programs are guaranteed to behave correctly when compiled and run. VST directly implements the concurrent separation logic of Hobor et al.~\cite{oraclesematic}, and the invariant-style lock specifications are incorporated into its soundness proof at a fundamental level.

@@

A thread-safe version of this data structure should satisfy atomic specifications that closely correspond to the sequential specs\footnote{We also have functions to create and deallocate trees, but these functions need not be thread-safe, since the tree will not be shared when created and must not be shared when deallocated.}:
\begin{mathpar}
\langle m.\ \nodeboxrep_g\ p\ |\ \treerep_g\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep_g\ p\ |\ \treerep_g\ (m[k \mapsto v])\rangle

@@

\subsection{Atomicity with lock invariants}
The easiest way to synchronize a data structure is to add a lock that
each operation acquires at the start and releases at the end. These
\emph{coarse-grained} locking implementations are easy to verify with
invariant-style lock specs: the lock's invariant can contain the whole
data structure assertion (e.g. $\mathsf{BST}\:p\ t$), so 
%that whichever thread is currently executing an operation on the tree at
%$p$ has full ownership of the data structure. 
a thread always has full ownership of the data structure when
performing an operation on $p$.  If we only want to prove that the operation
is thread-safe, we might choose the lock invariant $\exists t.\
\mathsf{BST}\ p\ t$, asserting that we gain full ownership of the tree
in some state $t$ every time we acquire the lock. On the other hand,
if we want to prove specifications that describe changes to the data
structure---as in the atomic specification $\langle m.\ \treerep\
m\rangle\ \texttt{insert}(p, k, v)\ \langle \treerep\ (m[k \mapsto
v])\rangle$---we need ghost state to track the changes.

@@

When insertion encounters a node with key \lstinline{k}, it simply changes that node's value to \lstinline{value}. If it reaches a leaf, it allocates a new node with key \lstinline{k} and value \lstinline{v}, and inserts it at that position.
In both cases, we must show that we can atomically update the abstract state of the tree from $\treerep\ m$ to $\treerep (m[\texttt{k}\mapsto \texttt{v}])$, by showing that the update to the tree in memory corresponds to a local change in the abstract tree that inserts the new key-value pair. The proofs are split into a proof that the C code implements the local update, and a proof that the local update to a ghost-state tree implements the global update. In the first case, the local change is the insertion of $(\texttt{k}, \texttt{v})$ at an empty leaf whose range includes \lstinline{k}, and the addition of two new empty leaves as its children; in the second case, we simply change the value of an internal node whose key is already \lstinline{k}.

@@

\begin{figure}[t]%[htp]
\begin{subfigure}[t]{0.6\textwidth}
\begin{lstlisting}[language = C, basicstyle=\small\ttfamily, numbers=left]
void pushdown_left (tree_t *tgt){
  while(1) {
    tree *p = tgt->t;
    tree_t *rc = p->right;
    void *lq = rc->lock;
    acquire(rc->lock);
    tree *q = rc->t;
    if (q == NULL) {
      tree_t *lc = p->left;
      ... // move lc->t to tgt; deallocate p, lc, and rc
      release(tgt->lock);
      return;
    } else {
      turn_left(tgt, rc);
      tgt = q->left;
      release(tgt->lock);
    }
  }
}
\end{lstlisting} 
\end{subfigure}\qquad
\begin{subfigure}[t]{0.4\textwidth}
 \begin{lstlisting}[language = C, basicstyle=\small\ttfamily]

@@

Function \lstinline{pushdown_left} changes the structure of the tree in two ways. First, the rotation \lstinline{turn_left} (line 14) changes the ranges of the node to be deleted and its right child (e.g., nodes 40 and 55 in the first step of figure~\ref{delete-diagram}); this is a local change, as the ranges of all other nodes (even the children of the rearranged nodes) are preserved. This is not a linearization point, so the rotation must be considered not to change the abstract state of the tree. We accounted for this by quantifying over the abstract tree $t$ in the definition of $\treerep$; rearranging nodes without changing the key-value map still implements the same abstract state, and other threads cannot assume that the tree structure they have observed is still the current structure of the tree. Second, the final removal of the node (line 10) increases the ranges of the subtree below it; for example, in the last step of figure~\ref{delete-diagram}, the range of node 35 changes from $(30, 40)$ to $(30, 50)$ when the node is removed. This is the linearization point of the function. The change occurs while \lstinline{pushdown_left} holds only the lock of the deleted node, so the algebra for range ghost state must allow a node's range to increase at any time, even when the node's lock is held. Fortunately, \lstinline{insert} and \lstinline{lookup} only use ranges to assert that the target key is in the range of the current node, which is still true if the range increases.

@@

\begin{figure}[t]%[htb]

@@

\section{Using the Specifications}
\begin{figure}[t]%[htp]
\begin{subfigure}[t]{0.4\textwidth}
\begin{lstlisting}[language = C, basicstyle=\small\ttfamily, numbers=left]
void* update_tree(void* t){
  a = 1;
  b = 2;
  c = 3; 
  d = 4;
  insert(t, 1, &a);
  insert(t, 2, &b);
  insert(t, 1, &c);
  insert(t, 2, &d);
  return NULL;
}
\end{lstlisting} 
\end{subfigure}\qquad
\begin{subfigure}[t]{0.55\textwidth}
 \begin{lstlisting}[language = C, basicstyle=\small\ttfamily]
int retrieve_value(treebox t){
  int y = -1;
  do {
     searchResult = lookup (t,  2);
     if (searchResult != NULL) {
        y = *((int *) searchResult);
     }
  } while (y != 4);
  void* searchResult = lookup (t,  1);
  return *((int *) searchResult);
}
\end{lstlisting}
\end{subfigure}
\caption{Client code}
\label{bst-conc-client}
\end{figure}
To demonstrate that the complexities of our lock-based proofs are hidden from prospective users, we verified a simple client program for the binary search tree, shown in figure \ref{bst-conc-client}. A producer thread running the \lstinline{update_tree} function populates a tree, while a consumer thread running \lstinline{retrieve_value} waits until key 2 is associated with a pointer to the value 4, and then returns the value associated with key 1. This is a data-structure-level version of the message-passing idiom, and demonstrates one of the advantages of atomic specifications: we can use them to synchronize threads and pass ownership of resources via data structure operations. The invariant for the BST \lstinline{t} is a simple state machine with three states, depending on the contents of the abstract map $m$: 1) $m(2) = \texttt{NULL}$, 2) $m(2) \mapsto i$ for some integer $i \neq 4$; and 3) $m(2) \mapsto 4 \land m(1) \mapsto 3$. In this final state, the invariant also holds the resources to be transferred from the producer to the consumer. We also include ghost state to reflect the fact that there is exactly one producer thread (so the producer always knows exactly which values are in the tree) and one consumer thread (so it can retrieve the resources without worrying about interference). The proof of the client uses only the atomic specifications of the BST operations, and is not affected by the style of lock specification.

@@

VeriFast~\cite{verifast}, a separation-logic-based verifier for C and
Java, supports lock-based and atomic
concurrency~\cite{verifast-conc}, and has been used to verify a
hand-over-hand-locking linked list similar to that of Krishna et al.
The specifications of the lock operations and the list itself use
a precursor of TaDA's logical atomicity. VeriFast is not 
foundational, but its basic logic is verified
against the semantics of a toy language in Coq.
